---
title: "flat_full.Rmd for working package"
output: html_document
editor_options: 
  chunk_output_type: console
---

<!-- Run this 'development' chunk -->
<!-- Store every call to library() that you need to explore your functions -->

```{r development, include=FALSE}
library(testthat)
library(usethis)
library(purrr)
library(janitor)
library(readxl)
```

<!--
 You need to run the 'description' chunk in the '0-dev_history.Rmd' file before continuing your code there.

If it is the first time you use {fusen}, after 'description', you can directly run the last chunk of the present file with inflate() inside.
--> 

```{r development-load}
# Load already included functions if relevant
pkgload::load_all(export_all = FALSE)
```

# Include some data examples in your package

<!-- 
 Store your dataset in a directory named "inst/" at the root of your project.
 Use it for your tests in this Rmd thanks to `pkgload::load_all()` to make it available
and `system.file()` to read it in your examples.

- There already is a dataset in the "inst/" directory to be used in the examples below
-->

```{r development-dataset}
# Run all this chunk in the console directly
# There already is a dataset in the "inst/" directory
# Make the dataset file available to the current Rmd during development
pkgload::load_all(path = here::here(), export_all = FALSE)

# You will be able to read your example data file in each of your function examples and tests as follows - see chunks below
datafile <- system.file("Pedigree_advanced_clones_SSA.xlsx", package = "pbwrangler")
sheet_name <- readxl::excel_sheets(datafile)
df_c <- map2(datafile,sheet_name, read_excel) %>% 
  set_names(tolower(gsub("\\s+", "_", sheet_name))) %>%
    map(., janitor::clean_names)

adv_clones <- df_c[["adv_clones"]]
adv_clones
family_code <- df_c[["family_code"]]
family_code

save(adv_clones, file = file.path(here::here(), "data", paste0("adv_clones", ".rda")))
save(family_code, file = file.path(here::here(), "data", paste0("family_code", ".rda")))
```


# The first function of the package: Calculate the median of a vector

<!--
Create a chunk for the core of the function

- The chunk needs to be named `function` at least
- It contains the code of a documented function
- The chunk can also be named `function-my_median` to make it easily
findable in your Rmd
- Let the `@examples` part empty, and use the next `examples` chunk instead to present reproducible examples

After inflating the template

-  This function code will automatically be added in a new file in the "R/" directory
-->

```{r function-my_left}

#' Extract first n elements of a string
#'
#' @param x a character vector of any length
#' @param num an integer - the number of elements to extract
#'
#' @return a character vector
#' @export
#'
#' @examples
my_left <- function(x,num){
  if(!is.character(x)) stop("x should be a character")
  stringr::str_sub(x, end = num)
}
```

```{r function-my_right}

#' Extract last n elements of a string
#'
#'
#' @param x a character vector of any length
#' @param num an integer - the number of elements to extract
#'
#' @return a character vector
#' @export
#'
#' @examples
my_right <- function(x, num){
  if(!is.character(x)) stop("x should be a character")
  stringr::str_sub(x, start = stringr::str_length(x)-num+1)
}

```

<!--
Create a chunk with an example of use for your function

- The chunk needs to be named `examples` at least
- It contains working examples of your function
- The chunk is better be named `examples-my_median` to be handled
correctly when inflated as a vignette

After inflating the template

-  This example will automatically be added in the '@examples' part of our function above in the "R/" directory
- This example will automatically be added in the vignette created from this Rmd template
-->

```{r examples-my_left}
a <- c("asdf", "bnmkk")
my_left(a, num = 3)
```

```{r examples-my_right}
a <- c("asdf", "bnmkk")
my_right(a, num = 3)
```

<!--
Create a chunk with a test of use for your function

- The chunk needs to be named `tests` at least
- It contains working tests of your function
- The chunk is better be named `tests-my_median` to be handled
correctly when inflated as a vignette

After inflating the template

-  This test code will automatically be added in the "tests/testthat/" directory
-->

```{r tests-my_left}
test_that("my_left works properly and show error if needed", {
  expect_true(my_left("absd", 2) == "ab")
  expect_error(my_left(112,3))
})

# Test with your dataset in "inst/"
# datafile <- system.file("nyc_squirrels_sample.csv", package = "pbwrangler")
# nyc_squirrels <- read.csv(datafile, encoding = "UTF-8")
# # Apply test on my function
# test_that("my_median works properly with internal dataset", {
#   expect_equal(my_median(nyc_squirrels[,"hectare_squirrel_number"]), 3)
# })
```

# Calculate the mean of a vector

<!--
There can be other functions, examples and tests in your flat template.
Each of them will be inflated in a different file, provided that there is a level-1 or level-2 section title to separate from previous functions.
-->

<!-- ## Use sub-functions in the same chunk -->

# ```{r function-my_other_median}
# #' My Other median
# #'
# #' @param x Vector of Numeric values
# #' @inheritParams stats::median
# #'
# #' @return
# #' Median of vector x
# #' @export
# #'
# #' @examples
# my_other_median <- function(x, na.rm = TRUE) {
#   if (!is.numeric(x)) {stop("x should be numeric")}
#   sub_median(x, na.rm =na.rm)
# }
# 
# #' Core of the median not exported
# #' @param x Vector of Numeric values
# #' @inheritParams stats::median
# sub_median <- function(x, na.rm = TRUE) {
#   stats::median(x, na.rm)
# }
# ```



<!--
# There can be development actions

Create a chunk with 'development' actions

- The chunk needs to be named `development` or `dev`
- It contains functions that are used for package development only
- Note that you may want to store most of these functions in the 0-dev_history.Rmd file

These are only included in the present flat template file, their content will not be part of the package anywhere else.
-->

```{r development-inflate, eval=FALSE}
# Keep eval=FALSE to avoid infinite loop in case you hit the knit button
# Execute in the console directly
fusen::inflate(flat_file = "dev/data_doc.Rmd", vignette_name = "Get started", overwrite = T)
fusen::inflate(flat_file = "dev/flat_full.Rmd", vignette_name = "Get started", overwrite = T)

```


# Inflate your package

You're one inflate from paper to box.
Build your package from this very Rmd using `fusen::inflate()`

- Verify your `"DESCRIPTION"` file has been updated
- Verify your function is in `"R/"` directory
- Verify your test is in `"tests/testthat/"` directory
- Verify this Rmd appears in `"vignettes/"` directory
